---
title: "WAHA Polyphenols Study"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required packages
pacs <- c("tidyverse", "haven", "janitor", "readxl", "tableone")
sapply(pacs, require, character.only = TRUE)

# Read functions
source("functions.R")

# Read data ---------------------------------------------------------------
# Will be using only LLU data -- make sure to subset for id >= 5000

zipfile <- "./data/Dietary Polyphenol Lipid and Inflammation files.zip"

lipid_file <- "BDLab_BCNLLU1709lipids.sav"
inflm_file <- "WAHA_ BD_Cytokines_BCN_LLU Inflammation.sav"
diet_file  <- "waha-recalls-per-recalls-overlapping-foodgroups-with-pOH.csv"
body_file  <- "Table1Data.xlsx"

source("dataprep.R")

# lipid data --------------------------------------------------------------
# n = 369
# Select variables and rename
lipid2 <- lipid %>% 
  select(-contains("hb"), -total_cholesterol1, -triglycerides1) %>%
  rename(TC_0 = total_cholesterol,
         TC_2 = total_cholesterol2,
         Trig_0 = triglycerides,
         Trig_2 = triglycerides2,
         VLDL_0 = vldl_c_b,
         VLDL_2 = vldl_c_f,
         LDL_0 = ldl_c_b,
         LDL_2 = ldl_c_f,
         HDL_0 = hdl_c_b,
         HDL_2 = hdl_c_f)

# inflammatory marker data ------------------------------------------------
# n = 371
# Rename variables
inflm2 <- inflm %>% 
  rename(hsCRP_0 = high_sensitivity_crp,
         hsCRP_2 = high_sensitivity_crp2,
         IL1_0 = il_1ss_1,
         IL1_2 = il_1ss_2,
         IL6_0 = il_6_1,
         IL6_2 = il_6_2,
         TNFa_0 = tnf_a_1,
         TNFa_2 = tnf_a_2) %>% 
  select(-high_sensitivity_crp1)

# Merge lipid and inflammation data, exlucding missing group
# n = 356
lipinf <- lipid2 %>% 
  select(-assigned_group) %>% 
  inner_join(inflm2, by = "patient_id") %>%
  filter(!is.na(assigned_group)) %>% 
  mutate(group = factor(assigned_group, labels = c("Walnut", "Control")),
         group = relevel(group, ref = "Control"),
         gender = factor(gender, labels = c("F", "M"))) %>% 
  select(-center, -dropout_visit) %>% 
  select(patient_id, group, assigned_group, gender, age, dropout, everything())

# Anthropometric data -----------------------------------------------------
# n = 356
# Select variables
# BMI needs to be re-calculated
body2 <- body %>% 
  mutate(educ = factor(education),
         educ_yr = education_years,
         race = factor(race),
         BMI = weight_basal / (height_basal / 100) ^ 2) %>% 
  select(patient_id, race, educ_yr, educ, race, BMI)

# Merge to lipid/inflammation data
lipinf2 <- lipinf %>% 
  inner_join(body2, by = "patient_id")

# Diet recall data --------------------------------------------------------
# n.obs = 1246, n.subj = 334

# There is one ID not found in lipid/inflammation merge file
ids_no_lipid <- drec %>% 
  distinct(patient_id) %>% 
  anti_join(lipinf, by = "patient_id")

# There are 23 IDs not found in dietary recalls
ids_no_recall <- lipinf %>% 
  anti_join(distinct(drec, patient_id), by = "patient_id") %>% 
  select(patient_id)

# Remove subjects with no lipid
# Rename long names
# n.obs = 1242, n.subj = 333
drec2 <- drec %>% 
  anti_join(ids_no_lipid, by = "patient_id") %>% 
  rename(flavanols = flavanols_proanthocyanidins_np)

# Energy-adjust each recall day and then take the average
pps <- c("total_polyphenol", "total_flavonoids", "flavanols", "phenolic_acid", "lignin")
ea_pps <- paste0(pps, "_ea")
drec2[ea_pps] <- lapply(drec2[pps], kcal_adjust, kcal = drec2$energy_kcal)

dintake <- drec2 %>% 
  group_by(patient_id) %>% 
  summarize_at(c("energy_kcal", ea_pps), mean, na.rm = TRUE)

# Merge three files -------------------------------------------------------
# n = 333 after merging
pp_df <- lipinf2 %>% 
  inner_join(dintake, by = "patient_id")

```

## Datasets

* A zip file receved from RA: `Dietary Polyphenol Lipid and Inflammation files.zip`
  * Contains:
    * Lipid file in SPSS: `BDLab_BCNLLU1709lipids.sav`
    * Inflammation markers file in SPSS: `WAHA_ BD_Cytokines_BCN_LLU Inflammation.sav`
    * Anthropometrics in Excel: `Table1Data.xlsx`
    * Recall files in CSV: `waha-recalls-per-recalls-overlapping-foodgroups-with-pOH.csv`

### Lipid data

* Includes *n* = `r nrow(lipid)` subjects.
* Variables: total cholesterol, LDL, HDL, triglycerides (mg/dL) at baseline and year 2. HbA1c values are all missing. 

### Inflammatory marker data

* Includes *n* = `r nrow(inflm)` subjects.
* Variables: hsCRP (mg/dL), IL-1 (pg/ml), IL-6 (pg/ml), TNF-a (pg/ml) at baseline and year 2.

### Anthropometric data

* Includes *n* = `r nrow(body)` subjects.
* Variables: Group, age, gender, race, education, height, weight, BMI, etc.

### Dietary recall data

* Includes `r nrow(drec)` observations from *n* = `r n_distinct(drec$patient_id)` subjects. The number of recalls ranges from 1 to 6 recalls.
  * There is 1 subject who was not found in lipid and inflammatory marker data. This subject was excluded, yielding `r nrow(drec2)` observations from *n* = `r n_distinct(drec2$patient_id)` subjects.
  
* Variables: Total energy (kcal/day), total polyphenols, total flavonoids, flavanols, phenolic acid, lignin, etc.
* Dietary polyphenol variables were energy-adjusted each day using the residual method and then averaged for each subject.

## Analytic dataset

```{r include=FALSE}
# Complete cases only -----------------------------------------------------

lipid_vars <- c("TC_0", "TC_2", "LDL_0", "LDL_2", "HDL_0", "HDL_2", "Trig_0", "Trig_2")
infl_vars  <- c("hsCRP_0", "hsCRP_2", "IL1_0", "IL1_2", "IL6_0", "IL6_2", "TNFa_0", "TNFa_2")
vars       <- c("TC", "LDL", "HDL", "Trig", "hsCRP", "IL1", "IL6", "TNFa")

# Complete cases only: n = 300
pp_df_comp_id <- pp_df %>% 
  select(patient_id, group, all_of(lipid_vars), all_of(infl_vars)) %>% 
  na.omit() %>% 
  select(patient_id)

pp_df_comp <- pp_df %>% 
  semi_join(pp_df_comp_id, by = "patient_id") %>% 
  mutate(TC_change = TC_2 - TC_0,
         LDL_change = LDL_2 - LDL_0,
         HDL_change = HDL_2 - HDL_0,
         Trig_change = Trig_2 - Trig_0,
         hsCRP_change = hsCRP_2 - hsCRP_0,
         IL1_change = IL1_2 - IL1_0,
         IL6_change = IL6_2 - IL6_0,
         TNFa_change = TNFa_2 - TNFa_0) 
```

* All data files were inner-joined, producing *n* = `r nrow(pp_df)` subjects.
* For analysis, any subjects who has any missing on lipids and inflammatory markers were excluded. This resulted in *n* = `r nrow(pp_df_comp)` subjects.

## Descriptive table at baseline by treatment

* Medians (and IQR) were reported for HDL, triglycerides and all inflammatory marker variables, as their distributions were right-skewed. Mann-Whitney tests were used for these variables for comparisons between treatment (`test = nonnormal`)

```{r echo = FALSE}

# Descriptive table at baseline -------------------------------------------

table_vars <- c("gender", "age", "BMI", "TC_0", "LDL_0", "HDL_0", "Trig_0", "hsCRP_0", "IL1_0", "IL6_0", "TNFa_0")

pp_df_comp %>% 
  CreateTableOne(table_vars, strata = "group", data = .) %>% 
  print(showAllLevels = TRUE, nonnormal = c("HDL_0", "Trig_0", "hsCRP_0", "IL1_0", "IL6_0", "TNFa_0")) %>%  
  kableone()
```

## Descriptive table of dietary intake of polyphenols by treatment

* All polyphenol intake variables were energy-adjusted as mentioned above. All dietary polyphenol variabls were right-skewed.

```{r echo = FALSE}

pp_df_comp %>% 
  CreateTableOne(ea_pps, strata = "group", data = .) %>% 
  print(nonnormal = ea_pps) %>%
  kableone()
```

* Density plots of dietary polyphenol intakes were shown below.

```{r echo = FALSE, fig.width = 10, fig.height = 3}
# Denstiy plot by group
pp_df_comp %>% 
  select(all_of(ea_pps), group) %>% 
  pivot_longer(1:5, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = ea_pps)) %>% 
  ggplot(aes(x = value, fill = group)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free", ncol = 5) +
  theme(legend.position = "bottom")

```

## Descriptive table of lipid/inflammatory markers by group and year

* Mean lipid and inflammator marker variables by group and year (baseline and year 2):

```{r echo = FALSE}

# Transform to long format
pp_df_long <- pp_df_comp %>% 
  pivot_longer(c(all_of(lipid_vars), all_of(infl_vars)), names_pattern = "(.*)_(.*)", names_to = c(".value", "year")) %>% 
  mutate(year = as.numeric(year)) 

# Descriptive table of lipids and inflammatory markers by year and group
pp_df_long %>% 
  CreateTableOne(vars, strata = c("year", "group"), data = ., test = FALSE) %>% 
  kableone()
```

* Changes from baseline were calculated for each subject and then average changes were calculated. The walnut group tended to show more declines in IL-1, IL-6 and TNF-a.

```{r echo = FALSE}

pp_df_comp %>% 
  CreateTableOne(paste0(vars, "_change"), strata = "group", data =., test = FALSE) %>% 
  kableone()
```

