---
title: "WAHA Polyphenols Study"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required packages
pacs <- c("tidyverse", "haven", "janitor", "readxl", "tableone", "tables")
sapply(pacs, require, character.only = TRUE)

# Read functions
source("functions.R")

# Read data ---------------------------------------------------------------
# Will be using only LLU data -- make sure to subset for id >= 5000

zipfile <- "./data/Dietary Polyphenol Lipid and Inflammation files.zip"

lipid_file <- "BDLab_BCNLLU1709lipids.sav"
inflm_file <- "WAHA_ BD_Cytokines_BCN_LLU Inflammation.sav"
diet_file  <- "waha-recalls-per-recalls-overlapping-foodgroups-with-pOH.csv"
body_file  <- "Table1Data.xlsx"
urine_file <- "./data/Master Polyphenol Database with Creatinine Final 04112022.xlsx"

source("dataprep.R")

# lipid data --------------------------------------------------------------
# n = 369
# Select variables and rename
lipid2 <- lipid %>% 
  select(-contains("hb"), -total_cholesterol1, -triglycerides1) %>%
  rename(TC_0 = total_cholesterol,
         TC_2 = total_cholesterol2,
         Trig_0 = triglycerides,
         Trig_2 = triglycerides2,
         VLDL_0 = vldl_c_b,
         VLDL_2 = vldl_c_f,
         LDL_0 = ldl_c_b,
         LDL_2 = ldl_c_f,
         HDL_0 = hdl_c_b,
         HDL_2 = hdl_c_f)

# inflammatory marker data ------------------------------------------------
# n = 371
# Rename variables
inflm2 <- inflm %>% 
  rename(hsCRP_0 = high_sensitivity_crp,
         hsCRP_2 = high_sensitivity_crp2,
         IL1_0 = il_1ss_1,
         IL1_2 = il_1ss_2,
         IL6_0 = il_6_1,
         IL6_2 = il_6_2,
         TNFa_0 = tnf_a_1,
         TNFa_2 = tnf_a_2) %>% 
  select(-high_sensitivity_crp1)

# Merge lipid and inflammation data, exlucding missing group
# n = 356
lipinf <- lipid2 %>% 
  select(-assigned_group) %>% 
  inner_join(inflm2, by = "patient_id") %>%
  filter(!is.na(assigned_group)) %>% 
  mutate(group = factor(assigned_group, labels = c("Walnut", "Control")),
         group = relevel(group, ref = "Control"),
         gender = factor(gender, labels = c("F", "M"))) %>% 
  select(-center, -dropout_visit) %>% 
  select(patient_id, group, assigned_group, gender, age, dropout, everything())

# Anthropometric data -----------------------------------------------------
# n = 356
# Select variables
# BMI needs to be re-calculated
body2 <- body %>% 
  mutate(educ = factor(education),
         educ_yr = education_years,
         race = factor(race),
         BMI = weight_basal / (height_basal / 100) ^ 2) %>% 
  select(patient_id, race, educ_yr, educ, race, BMI)

# Merge to lipid/inflammation data
lipinf2 <- lipinf %>% 
  inner_join(body2, by = "patient_id")

# Diet recall data --------------------------------------------------------
# n.obs = 1246, n.subj = 334

# There is one ID not found in lipid/inflammation merge file
ids_no_lipid <- drec %>% 
  distinct(patient_id) %>% 
  anti_join(lipinf, by = "patient_id")

# There are 23 IDs not found in dietary recalls
ids_no_recall <- lipinf %>% 
  anti_join(distinct(drec, patient_id), by = "patient_id") %>% 
  select(patient_id)

# Remove subjects with no lipid
# Rename long names
# n.obs = 1242, n.subj = 333
drec2 <- drec %>% 
  anti_join(ids_no_lipid, by = "patient_id") %>% 
  rename(flavanols = flavanols_proanthocyanidins_np)

# Energy-adjust each recall day and then take the average
pps <- c("total_polyphenol", "total_flavonoids", "flavanols", "phenolic_acid", "lignin")
ea_pps <- paste0(pps, "_ea")
drec2[ea_pps] <- lapply(drec2[pps], kcal_adjust, kcal = drec2$energy_kcal)

fg <- drec2 %>% 
  select(contains("_total_polyphenol"), 
         contains("_total_flavonoid"),
         contains("_flavanol"),
         contains("_phenolic_acid"),
         contains("_lignan")) %>% 
  names()

dintake <- drec2 %>% 
  group_by(patient_id) %>% 
  summarize_at(c("energy_kcal", ea_pps, fg), mean, na.rm = TRUE)

# Merge three files -------------------------------------------------------
# n = 333 after merging
pp_df <- lipinf2 %>% 
  inner_join(dintake, by = "patient_id")

```

## Datasets

* A zip file receved from RA: `Dietary Polyphenol Lipid and Inflammation files.zip`
  * Contains:
    * Lipid file in SPSS: `BDLab_BCNLLU1709lipids.sav`
    * Inflammation markers file in SPSS: `WAHA_ BD_Cytokines_BCN_LLU Inflammation.sav`
    * Anthropometrics in Excel: `Table1Data.xlsx`
    * Recall files in CSV: `waha-recalls-per-recalls-overlapping-foodgroups-with-pOH.csv`

### Lipid data

* Includes *n* = `r nrow(lipid)` subjects.
* Variables: total cholesterol, LDL, HDL, triglycerides (mg/dL) at baseline and year 2. HbA1c values are all missing. 

### Inflammatory marker data

* Includes *n* = `r nrow(inflm)` subjects.
* Variables: hsCRP (mg/dL), IL-1 (pg/ml), IL-6 (pg/ml), TNF-a (pg/ml) at baseline and year 2.

### Anthropometric data

* Includes *n* = `r nrow(body)` subjects.
* Variables: Group, age, gender, race, education, height, weight, BMI, etc.

### Dietary recall data

* Includes `r nrow(drec)` observations from *n* = `r n_distinct(drec$patient_id)` subjects. The number of recalls ranges from 1 to 6 recalls.
  * There is 1 subject who was not found in lipid and inflammatory marker data. This subject was excluded, yielding `r nrow(drec2)` observations from *n* = `r n_distinct(drec2$patient_id)` subjects.
  
* Variables: Total energy (kcal/day), total polyphenols, total flavonoids, flavanols, phenolic acid, lignin, etc.
* Dietary polyphenol variables were energy-adjusted each day using the residual method and then averaged for each subject.

## Analytic dataset

```{r include=FALSE}
# Complete cases only -----------------------------------------------------

lipid_vars <- c("TC_0", "TC_2", "LDL_0", "LDL_2", "HDL_0", "HDL_2", "Trig_0", "Trig_2")
infl_vars  <- c("hsCRP_0", "hsCRP_2", "IL1_0", "IL1_2", "IL6_0", "IL6_2", "TNFa_0", "TNFa_2")
vars       <- c("TC", "LDL", "HDL", "Trig", "hsCRP", "IL1", "IL6", "TNFa")

# Complete cases only: n = 300
pp_df_comp_id <- pp_df %>% 
  select(patient_id, group, all_of(lipid_vars), all_of(infl_vars)) %>% 
  na.omit() %>% 
  select(patient_id)

white_levels <- c(rep("Non-White", 4), "White")
pp_df_comp <- pp_df %>% 
  semi_join(pp_df_comp_id, by = "patient_id") %>% 
  mutate(TC_change = TC_2 - TC_0,
         LDL_change = LDL_2 - LDL_0,
         HDL_change = HDL_2 - HDL_0,
         Trig_change = Trig_2 - Trig_0,
         hsCRP_change = hsCRP_2 - hsCRP_0,
         IL1_change = IL1_2 - IL1_0,
         IL6_change = IL6_2 - IL6_0,
         TNFa_change = TNFa_2 - TNFa_0,
         Race2 = factor(white_levels[race]),
         Race2 = relevel(Race2, "White"),
         total_polyphenol_ea_100 = total_flavonoids_ea / 100) 

```

* All data files were inner-joined, producing *n* = `r nrow(pp_df)` subjects.
* For analysis, any subjects who has any missing on lipids and inflammatory markers were excluded. This resulted in *n* = `r nrow(pp_df_comp)` subjects.

## Descriptive table at baseline by treatment

* Medians (and IQR) were reported for HDL, triglycerides and all inflammatory marker variables, as their distributions were right-skewed. Mann-Whitney tests were used for these variables for comparisons between treatment (`test = nonnormal`)

```{r echo = FALSE}

# Descriptive table at baseline -------------------------------------------

table_vars <- c("gender", "Race2", "age", "BMI", "TC_0", "LDL_0", "HDL_0", "Trig_0", "hsCRP_0", "IL1_0", "IL6_0", "TNFa_0")

pp_df_comp %>% 
  CreateTableOne(table_vars, strata = "group", data = .) %>% 
  print(showAllLevels = TRUE, nonnormal = c("HDL_0", "Trig_0", "hsCRP_0", "IL1_0", "IL6_0", "TNFa_0")) %>%  
  kableone()
```

* Inflammatory variables appear to have very large outliers. Check with RA.

```{r echo = FALSE, fig.width = 10, fig.height = 3}
# Note outliers on inflammatory markers
pp_df_comp %>% 
  select(hsCRP_0, IL1_0, IL6_0, TNFa_0) %>% 
  pivot_longer(hsCRP_0:TNFa_0, names_to = "variable", values_to = "value") %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free", ncol = 4)

```

## Descriptive table of dietary intake of polyphenols by treatment

* All polyphenol intake variables were energy-adjusted as mentioned above. All dietary polyphenol variabls were right-skewed.

```{r echo = FALSE}

pp_df_comp %>% 
  CreateTableOne(ea_pps, strata = "group", data = .) %>% 
  print(nonnormal = ea_pps) %>%
  kableone()
```

* Density plots of dietary polyphenol intakes were shown below.

```{r echo = FALSE, fig.width = 10, fig.height = 3}
# Denstiy plot by group
pp_df_comp %>% 
  select(all_of(ea_pps), group) %>% 
  pivot_longer(1:5, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = ea_pps)) %>% 
  ggplot(aes(x = value, fill = group)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free", ncol = 5) +
  theme(legend.position = "bottom")

```

## Descriptive table of lipid/inflammatory markers by group and year

* Mean lipid and inflammator marker variables by group and year (baseline and year 2). 
  * Note that mean baseline values were higher in IL-1 and IL-6 for the walnut group, but their medians were actually lower (see the descriptive table at baseline above). 

```{r echo = FALSE}

# Transform to long format
pp_df_long <- pp_df_comp %>% 
  pivot_longer(c(all_of(lipid_vars), all_of(infl_vars)), names_pattern = "(.*)_(.*)", names_to = c(".value", "year")) %>% 
  mutate(year = as.numeric(year)) 

# Descriptive table of lipids and inflammatory markers by year and group
pp_df_long %>% 
  CreateTableOne(vars, strata = c("year", "group"), data = ., test = FALSE) %>% 
  kableone()
```

* Changes from baseline were calculated for each subject and then average changes were calculated. The walnut group tended to show more declines in IL-1, IL-6 and TNF-a.

```{r echo = FALSE}

pp_df_comp %>% 
  CreateTableOne(paste0(vars, "_change"), strata = "group", data =., test = FALSE) %>% 
  kableone()
```

## Polyphenol intake by food group

* Means and SDs of polyphenol intake by food group.
  * Note that "Misc" group have all zero.

```{r echo = FALSE}

Mean <- function(x) mean(x, na.rm = TRUE)
SD <- function(x) sd(x, na.rm = TRUE)

options(scipen = 999)
pp_df_comp %>% 
  select(all_of(fg)) %>% 
  pivot_longer(1:55, names_to = "variable", values_to = "value") %>% 
  mutate(
    pp = case_when(
      grepl("_total_polyphenol", variable) == TRUE ~ "Total_polyphenol",
      grepl("_total_flavonoid",  variable) == TRUE ~ "Total_flavonoids",
      grepl("_flavanol", variable) == TRUE ~ "Flavanol",
      grepl("_phenolic_acid", variable) == TRUE ~ "Phenolic_acid",
      grepl("_lignan", variable) == TRUE ~ "Lignan"),
    pp = factor(pp, levels = c("Total_polyphenol", "Total_flavonoids", "Flavanol", "Phenolic_acid", "Lignan")),
    fg = case_when(
      grepl("animal", variable) == TRUE ~ "Animal protein",
      grepl("beverage", variable) == TRUE ~ "Beverage",
      grepl("chocolate", variable) == TRUE ~ "Chocolates",
      grepl("fats_and_oil", variable) == TRUE ~ "Fat/oil",
      grepl("fruit", variable) == TRUE ~ "Fruits",
      grepl("grains", variable) == TRUE ~ "Grains",
      grepl("legume", variable) == TRUE ~ "Legumes",
      grepl("misc", variable) == TRUE ~ "Misc",
      grepl("nuts", variable) == TRUE ~ "Nuts/seeds",
      grepl("spices", variable) == TRUE ~ "Spices",
      grepl("vegetable", variable) == TRUE ~ "Vegetables"),
    fg = factor(fg))  %>% 
  tabular(Heading("Food group") * fg ~ Heading() * value * Heading() * pp * Format(digits = 1) * (Mean + SD), data = .)
options(scipen = 0)
```

## Association b/w dietary polyphenol and the change in lipids

``` {r include = FALSE}
# Specify covariates
covar <- c("age", "gender", "BMI", "lipid_lowering")

# Function for ANCOVA model
# ancova_mod <- function(data, yvar, basevar){
#   covar <- paste(covar, collapse = " + ")
#   fm <- formula(paste0(yvar, "~ total_polyphenol_ea + ", basevar, " + ", covar))
#   print(fm)
#   mod <- data %>% lm(fm, data = .)
#   print(ggResidpanel::resid_panel(mod, plots = "all"))
#   return(summary(mod))
# }

# Function for ANCOVA model
ancova_mod <- function(data, yvar, basevar){
  covar <- paste(covar, collapse = " + ")
  fm <- formula(paste0(yvar, "~ total_polyphenol_ea_100 + ", basevar, " + ", covar))
  cat(paste("Model: ", deparse1(fm)))
  data %>% 
    lm(fm, data = .) %>% 
    broom::tidy() %>% 
    mutate(p.value = scales::pvalue(p.value)) %>%
    knitr::kable(col.names = c("Predictor", "Beta", "SE", "t", "p-value"),
                 digits = c(0, 3, 3, 2, 3))
}

```

* For each of lipids (total cholesterol and LDL), a linear models were fitted using the change (year 2 - baseline) as the dependent variable and the dietary total polyphenols (energy-adjusted) as the main independent variable of interest. 
  * The unit of dietary total polyphenols was changed to 100 mg/day, due to small beta estimates.
  * The model adjusted for the baseline value of the lipid being modeled, as well as: gender (female as reference), age, BMI and the use of lipid lowering medications (yes/no, no as reference) at the baseline.

### Change in TC vs total polyphenol

* A scatterplot indicated that there is no association between the change in TC and dietary polyphenols.

```{r echo = FALSE, fig.width = 5}
pp_df_comp %>% ggp_scatter(TC_change, "Changes in total cholesterol (Year 2 - Baseline, mg/dL")

```

* There was no significant association between the change in TC and dietary polyphenol intake:

```{r echo = FALSE}
pp_df_comp %>% ancova_mod("TC_change", "TC_0")

```

### Change in LDL vs total polyphenol

* A scatterplot indicated that there is no association between the change in TC and dietary polyphenols.

```{r echo = FALSE, fig.width = 5}
pp_df_comp %>% ggp_scatter(LDL_change, "Changes in LDL (Year 2 - Baseline, mg/dL")

```

* There was no significant association between the change in LDL and dietary polyphenol intake:

```{r echo = FALSE}
pp_df_comp %>% ancova_mod("LDL_change", "LDL_0")

```

## Association b/w dietary polyphenol and lipids using mixed models

* To examine if there are any associations between dietary total polyphenols (energy-adjusted) and blood lipids (total cholesterol and LDL), scatterplots were produced. Lowess smoothed curves were fitted for for both years of measurements (Year 0 and Year 2).

* Scatterplots of TC and LDL (y-axis) against dietary total polyphenols (x-axis):
  * Note that the x-axis is on the log scale

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10}
vars <- c("total_polyphenol_ea", "TC", "LDL", "hsCRP", "IL1", "IL6", "TNFa")

# Scatterplots for TC and LDL
pp_df_long %>% 
  select(all_of(vars[1:3]), year) %>% 
  pivot_longer(TC:LDL, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = vars[-1]),
         year = factor(year)) %>% 
  ggplot(aes(x = total_polyphenol_ea, y = value, color = year)) +
  geom_point() +
  geom_smooth(span = 1) +
  scale_x_log10() +
  labs(x = "Energy-adjusted total polyphenols intake") +
  facet_wrap(~ variable, scales = "free")

```

* To see if the association between dietary total polyphenols and lipids is significant or not, mixed models were fitted using lipids (TC or LDL) as the dependent variable and log of energy-adjusted dietary total polyphenols as the main independent variable. The models included: year (0 or 2), its interaction with dietary polyphenol, age, gender and BMI as fixed-effects terms and subjects as a random-effects term.
  * From the mixed model, an estimated slope associated with log dietary polyphenol was calculated for each year.

* Estimated slope for dietary total polyphenol on TC (1st table below) and LDL (2nd table) by year:
  * There was no significant association with TC or LDL at baseline or Year 2

``` {r include = FALSE}
# Mixed model analysis
library(lme4); library(emmeans); library(lmerTest)
```

``` {r echo = FALSE}
# Mixed models for TC and LDL
mod1a <- lmer(TC ~ log(total_polyphenol_ea) * year + age + gender + BMI + (1|patient_id), data = pp_df_long)
mod1b <- update(mod1a, LDL ~ .)

mod1 <- list(mod1a, mod1b) 
names(mod1) <- c("TC", "LDL")
mod1 %>% map(function(x) test(emtrends(x, "year", var = "log(total_polyphenol_ea)")))
```

## Association b/w dietary polyphenol and the change in inflammatory markers

* For each of inflammatory markers (hsCRP, IL-1, IL-6, TNF-a), a linear models were fitted using the change (year 2 - baseline) as the dependent variable and the dietary total polyphenols (energy-adjusted) as the main independent variable of interest. The model adjusted for the baseline value of the inflammatory marker being modeld, as well as: gender (female as reference), age, BMI and the use of lipid lowering medications (yes/no, no as reference) at the baseline.

### Change in hsCRP vs total polyphenol

* A scatterplot showed there are several outliers in change in hsCRP. Check with RA.

```{r echo = FALSE, fig.width = 5}
pp_df_comp %>% ggp_scatter(hsCRP_change, "Changes in hsCRP (Year 2 - Baseline, mg/dL)")

```

* There was no significant association between the change in LDL and dietary polyphenol intake:

```{r echo = FALSE}
pp_df_comp %>% ancova_mod("hsCRP_change", "hsCRP_0")

```

### Change in IL-1 vs total polyphenol

* A scatterplot showed there are several outliers in change in IL-1. Check with RA.

```{r echo = FALSE, fig.width = 5}
pp_df_comp %>% ggp_scatter(IL1_change, "Changes in IL-1 (Year 2 - Baseline, pg/ml)")

```

* There was no significant association between the change in IL-1 and dietary polyphenol intake:

```{r echo = FALSE}
pp_df_comp %>% ancova_mod("IL1_change", "IL1_0")

```

### Change in IL-6 vs total polyphenol

* A scatterplot showed there are several outliers in change in IL-6. Check with RA.

```{r echo = FALSE, fig.width = 5}
pp_df_comp %>% ggp_scatter(IL6_change, "Changes in IL-6 (Year 2 - Baseline, pg/ml)")

```

* There was no significant association between the change in IL-6 and dietary polyphenol intake:

```{r echo = FALSE}
pp_df_comp %>% ancova_mod("IL6_change", "IL6_0")

```

### Change in TNF-a vs total polyphenol

* A scatterplot indicated that there is no association between the change in TNF-a and dietary polyphenols.

```{r echo = FALSE, fig.width = 5}
pp_df_comp %>% ggp_scatter(TNFa_change, "Changes in TNFa (Year 2 - Baseline, pg/ml)")

```

* There was no significant association between the change in TNF-a and dietary polyphenol intake:

```{r echo = FALSE}
pp_df_comp %>% ancova_mod("TNFa_change", "TNFa_0")

```

## Association b/w dietary polyphenol and inflammatory markers using mixed models

* To examine if there are any associations between dietary total polyphenols (energy-adjusted) and inflammatory markers (hsCRP, IL-1, IL-6, TNFa), scatterplots were produced. Lowess smoothed curves were fitted for for both years of measurements (Year 0 and Year 2).

* Scatterplots against dietary total polyphenols (x-axis):
  * Note that the both x- and y-axis are on the log scale

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 10}
# Scatterplots for inflammatory markers
pp_df_long %>% 
  select(all_of(vars[c(1, 4:7)]), year) %>% 
  pivot_longer(hsCRP:TNFa, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = vars[4:7]),
         year = factor(year)) %>% 
  ggplot(aes(x = total_polyphenol_ea, y = value, color = year)) +
  geom_point() +
  geom_smooth(span = 1) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Energy-adjusted total polyphenols intake") +
  facet_wrap(~ variable, scales = "free", ncol = 2)
```

* To see if the association between dietary total polyphenols and inflammatory markers is significant or not, mixed models were fitted using inflammatory markers as the dependent variable and log of energy-adjusted dietary total polyphenols as the main independent variable. The models included: year (0 or 2), its interaction with dietary polyphenol, age, gender and BMI as fixed-effects terms and subjects as a random-effects term.
  * From the mixed model, an estimated slope associated with log dietary polyphenol was calculated for each year.

* Estimated slope for dietary total polyphenol on each inflammatory marker by year:
  * There was no significant association with any of inflammatory markers at baseline or Year 2

``` {r echo = FALSE}
# Mixed models for inflammatory markers
mod2a <- lmer(log(hsCRP) ~ log(total_polyphenol_ea) * year + age + gender + BMI + (1|patient_id), data = pp_df_long)
mod2b <- update(mod2a, log(IL1) ~.)
mod2c <- update(mod2a, log(IL6) ~.)
mod2d <- update(mod2a, log(TNFa) ~.)

mod2 <- list(mod2a, mod2b, mod2c, mod2d) 
names(mod2) <- c("log(hsCRP)", "log(IL1)", "log(IL6)", "log(TNFa)")
mod2 %>% map(function(x) test(emtrends(x, "year", var = "log(total_polyphenol_ea)")))
```

## Analysis on urinary polyphenol

### Descriptive analysis by treatment and time

* Means (SD) by treatment and time are shown below for urine polyphenol (`ur_tot_pp`) and urine polyphenol / creatinine (`ur_tot_pp_cr`):

```{r echo = FALSE}
# When merged with lipid/diet wide format data, 295 unique subjects
urine_comp <- urine %>% 
  rename(ur_tot_pp = polyphenol_yield_correction_factor_2,
         ur_tot_pp_cr = final_polyphenol_yield_mg_g_creatinine) %>% 
  inner_join(pp_df_comp, by = "patient_id")

# Descriptive stats by diet and time
urine_comp %>% 
  tabular(Heading("Year") * time ~ Heading("") * group * (ur_tot_pp + ur_tot_pp_cr) * (Mean + SD), data = .)
```

### Comparisons of urine polyphenol between treatment groups over time

* To compare urine polyphenol between two treatment groups over time, mixed models were fitted for both variables (`ur_tot_pp` and `ur_tot_pp_cr`). The mixed model included treatment, time and treatment x time interaction as fixed-effects terms and subjects as a random-effects term.

  * From the mixed model, marginal means were estimated for the two treatment groups at each time point and then compared, by testing for the simple effects of treatment at each time point.

* **Urine total polyphenol**: Estimated marginal means by treatment group at each time point

``` {r echo = FALSE}
# No creatinine adjustment
mod1 <- lmer(ur_tot_pp ~ group * time + (1|patient_id), data = urine_comp)
mod1_emm <- emmeans(mod1, ~ group| time)
mod1_emm %>% knitr::kable() 
```

* Comparisons between treatments by time:
  * There were no significant differences at any time point:

``` {r echo = FALSE}
pairs(mod1_emm, reverse = TRUE) %>% knitr::kable()
```

* **Urine total polyphenol adjusted for creatinine**: Estimated marginal means by treatment group at each time point

``` {r echo = FALSE}
# With creatinine adjustment
mod2 <- lmer(ur_tot_pp_cr ~ group * time + (1|patient_id), data = urine_comp)
mod2_emm <- emmeans(mod2, ~ group| time)
mod2_emm %>% knitr::kable()
```

* Comparisons between treatments by time:
  * There were no significant differences at any time point:

``` {r echo = FALSE}
pairs(mod2_emm, reverse = TRUE) %>% knitr::kable()
```

### Association between dietary polyphenol intake and urine polyphenol excretion

* To examine if there are any associations between dietary polyphenol intake and urine polyphenol at Year 2, scatterplots were produced. Dietary intakes (total, flavonoids, flavanols, phenonic acids) were energy-adjusted as described above. Note that in the scatterplots below the x-axis is on the log-scale. A lowess smoothed curve was fitted for each plot.

* Scatterplots against **urine total polyphenol**:

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 3}
# Scatterplots
pp_names <- urine_comp %>% select(ends_with("_ea")) %>% select(-lignin_ea) %>% names()

urine_comp_y2 <- urine_comp %>% filter(time == 2) %>% select(-lignin_ea)
ggp_df <- urine_comp_y2 %>% 
  pivot_longer(ends_with("_ea"), names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, labels = pp_names))

ggp_df %>% 
  ggplot(aes(x = value, y = ur_tot_pp)) +
  geom_point() +
  geom_smooth(method = "loess", span = 1) + 
  scale_x_log10() +
  facet_grid(~ variable, scales = "free_x") +
  labs(y = "Urinary total polyphenol")
```

* Scatterplots against **urine total polyphenol adjusted for creatinine**:

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 3}
ggp_df %>% 
  ggplot(aes(x = value, y = ur_tot_pp_cr)) +
  geom_point() +
  geom_smooth(method = "loess", span = 1) + 
  scale_x_log10() +
  facet_grid(~ variable, scales = "free_x") +
  labs(y = "Urinary total polyphenol / creatinine")
```

* A linear model was fitted for each combination of urine polyphenol (as a dependent variable) and log dietary polyphenol (an independent variable), while adjusting for age, gender and BMI.

* Linear models for **urine total polyphenol**:
  * There was a significantly **negative** association between urine total polyphenol and the log of dietary total flaonoids (p = 0.0208). There were no significant associations with any other dietary polyphenols.

``` {r echo = FALSE}
# Model
mod1a <- lm(ur_tot_pp ~ log(total_polyphenol_ea) + age + gender + BMI, data = urine_comp_y2)
mod1b <- lm(ur_tot_pp ~ log(total_flavonoids_ea) + age + gender + BMI, data = urine_comp_y2)
mod1c <- lm(ur_tot_pp ~ log(flavanols_ea + 1) + age + gender + BMI, data = urine_comp_y2)
mod1d <- lm(ur_tot_pp ~ log(phenolic_acid_ea) + age + gender + BMI, data = urine_comp_y2)

mod1 <- list(mod1a, mod1b, mod1c, mod1d)
names(mod1) <- paste0("log_", pp_names)
mod1 %>% map(function(x) summary(x)$coef) %>% map(round, 4)

```

* Linear models for **urine total polyphenol / creatinine**:
  * There was a significantly **negative** association between urine total polyphenol/creatinine and the log of dietary total flaonoids (p = 0.0456). There were no significant associations with any other dietary polyphenols.

``` {r echo = FALSE}
mod2a <- lm(ur_tot_pp_cr ~ log(total_polyphenol_ea) + age + gender + BMI, data = urine_comp_y2)
mod2b <- lm(ur_tot_pp_cr ~ log(total_flavonoids_ea) + age + gender + BMI, data = urine_comp_y2)
mod2c <- lm(ur_tot_pp_cr ~ log(flavanols_ea + 1) + age + gender + BMI, data = urine_comp_y2)
mod2d <- lm(ur_tot_pp_cr ~ log(phenolic_acid_ea) + age + gender + BMI, data = urine_comp_y2)

mod2 <- list(mod2a, mod2b, mod2c, mod2d)
names(mod2) <- paste0("log_", pp_names)
mod2 %>% map(function(x) summary(x)$coef) %>% map(round, 4)
```

### Association between urine polyphenol and blood lipids

* To examine if there are any associations between urine polyphenol and blood lipids (total cholesterol and LDL), scatterplots were produced. Lowess smoothed curves were fitted for for both years of measurements (Year 0 and Year 2).

* Scatterplots of TC and LDL (y-axis) against **urine total polyphenol** (x-axis):

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10}
# Urine polyphenol and blood lipids (TC, LDL)
urine_comp_long <- urine %>% 
  rename(ur_tot_pp = polyphenol_yield_correction_factor_2,
         ur_tot_pp_cr = final_polyphenol_yield_mg_g_creatinine) %>%
  mutate(year = as.numeric(time) - 1) %>% 
  inner_join(pp_df_long, by = c("patient_id", "year")) %>%
  mutate(year = factor(year)) %>% 
  arrange(patient_id, year)

# Scatterplots
urine_comp_long %>% 
  pivot_longer(TC:LDL, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("TC", "LDL"))) %>% 
  ggplot(aes(x = ur_tot_pp, y = value, color = year)) +
  geom_point() +
  geom_smooth(span = 1) +
  facet_grid(~ variable, scales = "free")
```

* Scatterplots of TC and LDL (y-axis) against **urine total polyphenol / creatinine** (x-axis):

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10}
urine_comp_long %>% 
  pivot_longer(TC:LDL, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("TC", "LDL"))) %>% 
  ggplot(aes(x = ur_tot_pp_cr, y = value, color = year)) +
  geom_point() +
  geom_smooth(span = 1) +
  facet_grid(~ variable, scales = "free")
```

* To see if the association between urine polyphenol and lipids is significant or not, mixed models were fitted using lipids (TC or LDL) as the dependent variable and urine polyphenol (with or without creatinine adjustment) as the main independent variable. The models included: year (0 or 2), its interaction with urine polyphenol, age, gender and BMI as fixed-effects terms and subjects as a random-effects term.
  * From the mixed model, an estimated slope associated with urine polyphenol (divided by 100) was calculated for each year.  

* Estimated slope for **urine total polyphenol** on TC (1st table below) and LDL (2nd table) by year:
  * There was a significant negative association with TC at Year 2 (p = 0.048).
  * No significant association was found with LDL.
  
``` {r include = FALSE}
# Mixed model analysis
library(lme4); library(emmeans); library(lmerTest)
```

``` {r echo = FALSE}
# No creatinine adjustment
mod1a <- lmer(TC ~ I(ur_tot_pp/100) * year + age + gender + BMI + (1|patient_id), data = urine_comp_long)
mod1b <- update(mod1a, LDL ~ .)

mod1 <- list(mod1a, mod1b) 
names(mod1) <- c("TC", "LDL")
mod1 %>% map(function(x) test(emtrends(x, "year", var = "I(ur_tot_pp)/100")))
```

* Estimated slope for **urine total polyphenol / creatinine** on TC (1st table below) and LDL (2nd table) by year:
  * There was a significant negative association with TC at both Year 0 (p = 0.024) and Year 2 (p = 0.0028).
  * No significant association was found with LDL.
  
``` {r echo = FALSE}
# With creatinine adjustment
mod2a <- lmer(TC ~ I(ur_tot_pp_cr/100) * year + age + gender + BMI + (1|patient_id), data = urine_comp_long)
mod2b <- update(mod2a, LDL ~ .)

mod2 <- list(mod2a, mod2b)
names(mod2) <- c("TC", "LDL")
mod2 %>% map(function(x) test(emtrends(x, "year", var = "I(ur_tot_pp_cr)/100")))
```

### Association between urine polyphenol and inflammatory markers

* To examine if there are any associations between urine polyphenol and inflammatory markers (CRP, IL-1, IL-6, TNF-a), scatterplots were produced. lowess smoothed curves was fitted for for both years of measurements (Year 0 and Year 2).

* Scatterplots of inflammatory markers (y-axis) against **urine total polyphenol** (x-axis):

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 3}
# Scatterplots
urine_comp_long %>% 
  select(patient_id, year, hsCRP, IL1, IL6, TNFa, starts_with("ur_")) %>% 
  pivot_longer(hsCRP:TNFa, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("hsCRP", "IL1", "IL6", "TNFa"))) %>% 
  ggplot(aes(x = ur_tot_pp, y = value, color = year)) +
  geom_point() +
  geom_smooth(span = 1) +
  scale_y_log10() +
  facet_grid(~ variable, scales = "free")
```

* Scatterplots of inflammatory markers (y-axis) against **urine total polyphenol / creatinine** (x-axis):

``` {r echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 3}
urine_comp_long %>% 
  select(patient_id, year, hsCRP, IL1, IL6, TNFa, starts_with("ur_")) %>% 
  pivot_longer(hsCRP:TNFa, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("hsCRP", "IL1", "IL6", "TNFa"))) %>% 
  ggplot(aes(x = ur_tot_pp_cr, y = value, color = year)) +
  geom_point() +
  geom_smooth(span = 1) +
  scale_y_log10() +
  facet_grid(~ variable, scale = "free")
```

* To see if the association between urine polyphenol and inflammatory marker is significant or not, mixed models were fitted using each inflammatory marker as the dependent variable and urine polyphenol (with or without creatinine adjustment) as the main independent variable. The models included: year (0 or 2), its interaction with urine polyphenol, age, gender and BMI as fixed-effects terms and subjects as a random-effects term.
  * From the mixed model, an estimated slope associated with urine polyphenol (divided by 100) was calculated for each year.  

* Estimated slope for **urine total polyphenol** on inflammatory markers by year:
  * There were no significant association with any of inflmmatory markers.

``` {r echo = FALSE}
# No creatinine adjustment
mod1a <- lmer(log(hsCRP) ~ I(ur_tot_pp/100) * year + age + gender + BMI + (1|patient_id), data = urine_comp_long)
mod1b <- update(mod1a, log(IL1) ~ .)
mod1c <- update(mod1a, log(IL6) ~ .)
mod1d <- update(mod1a, log(TNFa) ~ .)

mod1 <- list(mod1a, mod1b, mod1c, mod1d)
names(mod1) <- c("log_hsCRP", "log_IL1", "log_IL6", "log_TNFa")
mod1 %>% map(function(x) test(emtrends(x, "year", var = "I(ur_tot_pp)/100")))

```  

* Estimated slope for **urine total polyphenol / creatinine** on inflammatory markers by year:
  * There were no significant association with any of inflmmatory markers.

``` {r echo = FALSE}

# With creatinine adjustment
mod2a <- lmer(log(hsCRP) ~ I(ur_tot_pp_cr/100) * year + age + gender + BMI + (1|patient_id), data = urine_comp_long)
mod2b <- update(mod2a, log(IL1) ~ .)
mod2c <- update(mod2a, log(IL6) ~ .)
mod2d <- update(mod2a, log(TNFa) ~ .)

mod2 <- list(mod2a, mod2b, mod2c, mod2d)
names(mod2) <- c("log_hsCRP", "log_IL1", "log_IL6", "log_TNFa")
mod2 %>% map(function(x) test(emtrends(x, "year", var = "I(ur_tot_pp_cr)/100")))
```  
